Deciphering enhancer rules underlying early human neuronal and neural crest development.

### 2025-04-08 manuscript.tex
S.A. comments

Genomes of complex multi-cellular organisms encode all cell types/states\cite{trapnellDefiningCellTypes2015, arendtOriginEvolutionCell2016, fleckWhatCellType2023} of that organism in specific non-coding elements named
enhancers\cite{davidsonRegulatoryGenomeGene2010}. The code words used by enhancers are
transcription factor binding sites (TFBS)\cite{davidsonRegulatoryGenomeGene2010} and the precise arrangement, copy
number and strength of individual TFBS define the enhancer and its cell type-specific activity\cite{dewinterModellingDesignTranscriptional2025}. Due to the soft syntax rules underlying the enhancer-code\cite{zeitlingerSevenMythsHow2020}, DNA sequence based deep learning methods have shown major advances in the ability to model and explain enhancers\cite{zeitlingerSevenMythsHow2020, dewinterModellingDesignTranscriptional2025}. For example, during immune cell differentiation\cite{maslovaDeepLearningImmune2020}, fly brain cell types\cite{janssensDecodingGeneRegulation2022}, liver cell types\cite{bravogonzalez-blasSinglecellSpatialMultiomics2024}, melanoma states\cite{minnoyeCrossspeciesAnalysisEnhancer2020, atakInterpretationAllelespecificChromatin2021}, during human brain development\cite{mannensChromatinAccessibilityHuman2024}, mammalian and chicken brain cell types\cite{heckerEnhancerdrivenCellType2025} and pluripotent cell stem cells\cite{avsecBaseresolutionModelsTranscriptionfactor2021}. Furthermore, it has been demonstrated recently that the code learned by these models is sufficient to design novel enhancers with cell type-specific activity\cite{taskiranCelltypedirectedDesignSynthetic2024, dealmeidaTargetedDesignSynthetic2024, gosaiMachineguidedDesignCelltypetargeting2024, yinIterativeDeepLearningdesign2024}.\par

Organoids are a reductionistic biological model system that can be used to interrogate how cell types are encoded in the
genome. Organoids derived from pluripotent stem cells are especially interesting in this context, given that they
demonstrate how a single genome can generate multiple distinct cell types\cite{mccauleyPluripotentStemCellderived2017,
takebeOrganoidsDesign2019, hoferEngineeringOrganoids2021}. Organoids can furthermore give insights into how genomes
interpret both cell intrinsic and extrinsic signals. For example, epigentic
modifications\cite{zenkSinglecellEpigenomicReconstruction2024}, transcription factor (TF)
perturbations\cite{fleckInferringPerturbingCell2023} or morphogen
signaling\cite{sanchis-callejaDecodingMorphogenPatterning2024}. Another benefit of organoids is that they can be used to
model biological systems that are otherwise - due to ethical and/or practical reasons - inaccessible for study. For
example, early human development.
One such system is early human development, the study of which is especially important for our understanding of human health and disease.
\par

@@

Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols and how they compare to their primary counterparts at a transcriptomic level in for example, brain organoids\cite{heIntegratedTranscriptomicCell2024} and endoderm-derived organoids\cite{xuIntegratedTranscriptomicCell2023}. However, such comparisons at the level of the enhancer-code are still limited\cite{zenkSinglecellEpigenomicReconstruction2024, fleckInferringPerturbingCell2023}. In this study, we will address this gap by determining the enhancer-codes of cell types/states generated by human neural tube organoids, a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021} and compare these codes to those of cell types/states of the head of a four weeks post conception human embryo.\par

During vertebrate development, both the neural tube and neural crest originate from the neural ectoderm. The neural tube
consists of proliferating neuronal progenitors that, through opposing gradients of SHH and BMP/Wnt signaling, have different identities along the dorso-ventral (DV) axis of the embryo. These early identities later give rise to distinct neuronal subtypes of the central nervous system\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020}. Neural crest cells originate from the neural plate border, through intermediate signals of BMP and Wnt\cite{simoes-costaEstablishingNeuralCrest2015}, and migrate throughout the body giving rise to a multitude of cell types including the peripheral nervous system, craniofacial skeleton and melanocytes\cite{simoes-costaEstablishingNeuralCrest2015}. The gene regulatory networks underlying both DV patterning of neuronal progenitors\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020} and neural crest induction, specification, migration and differentiation\cite{simoes-costaEstablishingNeuralCrest2015, candido-ferreiraMultilayeredTranscriptionalControl2023} are already well studied. Furthermore, studies using chromatin accessibility profiling methods at the bulk and single cell level, profiling enhancers encoding cell types and states of neuronal progenitors and neural crest cells are starting to emerge\cite{delasDevelopmentalCellFate2023, zhangCisregulatoryLogicIntegrating2024, huEndothelialRegulatoryCircuits2025} but a detailed descriptions of the enhancer-code of individual cell types are still limited.\par

@@

We made use of human neural tube organoids (hNTOs) to model early human neuronal and neural crest development. To this

@@

neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. In order to characterize the organoid-derived cell types to a greater level of detail at both

@@

Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) and in the
embryo only we could identify an endothelial, perivascular macrophage (PVM) / microglia (MIC) and otic placode
population (\textbf{Fig. 1c}).\par

@@

(\textit{SIX3}) or anterior spinal cord (\textit{HOXB3}) identity, we observed a gradient of anterior-to-posterior identities in the
embryo (REF): from anterior to posterior expression of \textit{SIX3}, \textit{EMX2}, \textit{FOXG1}, \textit{BARHL1}, \textit{EN1 }and \textit{HOXB3} (\textbf{Fig. 1g}).\par

@@

"$DeepNeuralTube_{organoid}$" and "$DeepNeuralTube_{embryo}$". The models use DNA sequence, of genomic regions, as input and

@@

train $DeepNeuralTube_{organoid}$) and embryo cells (to train $DeepNeuralTube_{embryo}$). To validate the ability of the models to

@@

(auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ (\textbf{Fig. 2b}; and

@@

enhancers with an auROC of 0.78 and 0.79 using resp. $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$

@@

Next, we wanted to assess the explainability of the models. Concretely we wanted to test whether the model learned known

@@

TF-MoDISco (\textbf{REF}) to find \textit{de novo} motifs used by the models. For both known motifs and \textit{de novo}

@@

(\textbf{Fig. 2e}) regions with the exception of cluster 13 which is organoid specific (AP1 motif; \textbf{Supplementary Table S2}) and cluster

@@

Thus, $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ use known-motifs as features to classify regions based

@@

$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ explain this region similarly (Pearson correlation of x) and

@@

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_2.png}
  \caption{
    \textbf{DeepNeuralTube: an accurate sequence-to-function model revealing neuro-developmental enhancer code.} \\
    \textbf{a}, Schematic. \textbf{b} receiver-operator curves for identifying neural tube (left) and facial mesenchyme the
    $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$. \textbf{c-e}, UMAP of TomTom [REF] motif-to-motif similarities
    using both \textit{de novo} motifs from DeepNeuralTube and known enriched motifs. Each dot is a motif. Motifs are
    colored by: cluster identity (\textbf{c}), model of origin (Continues next page)
  }
\end{figure}
\afterpage{\clearpage}

\begin{figure}[b]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) (\textbf{d}; only showing \textit{de novo} motifs in this case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
 \textbf{f}, Example representative motifs from clusters in b, showing motifs from motif enrichment (left) and found
    \textit{de novo} (right). \textbf{g}, contribution score for reference and mutant version of Vista enhancer hs111
    for $DeepNeuralTube_{organoid}$ (top two panels) and $DeepNeuralTube_{embryo}$ (bottom two panels) and enhancer-reporter
    assay for reference and mutant version of hs111 (right two panels) [REF]. \textbf{h}, contribution score for
    $DeepNeuralTube_{organoid}$ (top) and $DeepNeuralTube_{embryo}$ (bottom) of a genomic enhancer human neural crest
    enhancer (chr1:161200798-161201298; hg38) and enhancer-reporter assay in chicken embryo (HH4; right). Green is enhancer signal, pink is
    constitutive enhancer that serves as electroporation control.
  }
\end{figure}

\subsection*{
  Deciphering the enhancer code underlying dorso-ventral patterning of neuronal progenitors
}
Focusing on neuronal progenitors, we identified topics representing cells and regions of different dorso-ventral
identities in both the organoid and the embryo data (\textbf{Fig. 3a}). From ventral to dorsal this is topic 36, 38, 33,
54 and 48 in the organoid data and topic 34, 38, 79, 88 and 58 in the embryo data. As an example we first investigated
two enhancers of \textit{SHH} that are strongly conserved and specific to the floorplate [\textbf{REF}]; SFPE1 and SFPE2.
\textit{SHH} is most strongly expressed in the most ventral topic (33 and 34 resp. in the organoids and the embryo; \textbf{Fig. 3b}).
SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells
belonging to the ventral topic, in both the organoid and embryo data (\textbf{Fig. 3c}). Explaining the accessibility of both
enhancers using DeepNeuralTube reveals two FOX (one partial) and two RFX (one partial) binding sites in SFPE1
and two FOX, a ZEB and a TEAD binding site in SFPE2 (\textbf{Fig. 3d}).\par

To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we calculated
contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$. We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reoccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities. Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3e}). Both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns
underlying the different dorso-ventral states and the importance of the instances of these patterns is highly correlated across both
models (0.75 average Pearson correlation; \textbf{Fig. 3e}). For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD
and RFX instances is important (\textbf{Fig. 3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo cells (\textbf{Fig. 3g} and Supplementary Fig. S3). In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3e- g)}); PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3). For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), SOX instances are important which correlates with \textit{SOX2/3} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3). Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e} and Supplementary Fig. S3). Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig. 3h}). FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns seem to be present in separate regions (\textbf{Fig. 3h}).\par

It has been reported that FOXA1/2 is a pioneer factor (REF). Indeed, when centering on FOX instances in both the organoid and
embryo data we observe a strong footprint when looking at Tn5 cut-sites (\textbf{Fig. 3i}). We then
asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this
TF and regions, bound by FOXA2, that are specific to the floorplate. For this purpose, we downloaded FOXA2 ChIP-seq
peaks in HepG2 and A549 and intersected these with floorplate peaks containing FOX instances. Those floorplate peaks
that overlapped with both ChIP-seq datasets we classified as regions that are generally bound by FOXA2 and those that
did not we classified as floorplate specific. Next, we quantified both the number of FOX instances per region and the
affinity of each individual instance (making use of the FOX motif score calculated using
FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) hypothesizing that the overall affinity of regions that are generally
bound by FOXA2 is higher than floorplate specific regions either because these general regions have more instances or
the affinity of individual instances is higher. Indeed, FOXA2 target regions that are specific to the floorplate mostly
have a single FOX instance while FOXA2 target regions that are general tend to have more instances (\textbf{Fig. 3j}).
Furthermore, in case there is but a single FOX instance in a general FOXA2 target region the affinity of this instance tends
to be higher (\textbf{Fig. 3k}). Generating a \textit{de novo} motif from general and specific FOXA2 target regions that
have a single instance, we observe that FOX instances in regions that are general tend to be of the form
\textit{TGTTT\underline{AC}} while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).\par

@@

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38. SFPE2 and SFPE1 is highlighted. \textbf{d}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). \textbf{e}, Code table showing average contribution score of pattern instances for each dorso-ventral topic. Contribution scores for $DeepNeuralTube_{organoid}$ are not outlined. Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated. \textbf{f}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. \textbf{g}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. \textbf{h}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. \textbf{i}, Scaled average and normalized number of cut sites (black, top) and contribution score (red, top) across genomic regions with a FOXA2 instance and heatmap showing normalized number of cut sites (bottom) for individual regions with a FOXA2 instance. Signal is centered on the FOXA2 instance start position. \textbf{j}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIp-seq signal (general; bottom) and those regions that don't overlap (specific; top). \textbf{k}, Boxplot (ADD NUMEBRS) showing $-log_10(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grantFIMOScanningOccurrences2011,schreiberTangermemeBiologicalSequence2025} for general regions (bottom) and specific regions (top) stratified by the number of FOXA2 instance per region. \textbf{l}, \textit{de novo} motif for regions with a single FOXA2 instance that are specific (left, top) or general (right, top) and heatmap color coded by nucleotide identity for individual regions for both specific (left, bottom) and general (right, bottom) regions that have a single FOXA2 instance.
  }
\end{figure}

@@

Given that we observed cell states from different AP domains in the neuronal progenitors from the embryo (but not
organoid; see \textbf{Fig. 1}) we decided to study the gene regulatory code underlying this patterning further. We
identified seven topics corresponding to different AP domains, from anterior to posterior this is Topic 61, Topic 59,
Topic 31, Topic 62, Topic 70, Topic 52 and Topic 71 (\textbf{Fig. 4a}). We first asked if the AP-program is distinct
from the DV program. In other words, is AP patterning encoded in the same or in different genomic regions as DV patterning? For this purpose, we calculated the pairwise correlation between region-topic
probabilities of all AP and DV topics (\textbf{Fig. 4b}). From this it is clear that both AP and DV patterning as well
as each of their sub-domains are mostly encoded in distinct genomic regions, with the exception of AP Topic 59 and 31
that correlate with the region-topic probability of DV Topic 38 (which corresponds to cells expressing \textit{NKX2-1/2}; \textbf{Fig. 4b})\par

>>>>

Genomes of complex multi-cellular organisms encode all cell types/states\cite{trapnellDefiningCellTypes2015, arendtOriginEvolutionCell2016, fleckWhatCellType2023} of that organism in specific non-coding elements named
enhancers\cite{davidsonRegulatoryGenomeGene2010}. The code words used by enhancers are
transcription factor binding sites (TFBS)\cite{davidsonRegulatoryGenomeGene2010} and the arrangement, copy
number and strength of individual TFBS define the enhancer and its cell type-specific activity\cite{dewinterModellingDesignTranscriptional2025}. Due to the soft syntax rules underlying the enhancer-code\cite{zeitlingerSevenMythsHow2020}, DNA sequence based deep learning methods have shown major advances in the ability to model and explain enhancers\cite{zeitlingerSevenMythsHow2020, dewinterModellingDesignTranscriptional2025}. For example, immune cell differentiation\cite{maslovaDeepLearningImmune2020}, fly brain cell types\cite{janssensDecodingGeneRegulation2022}, liver cell types\cite{bravogonzalez-blasSinglecellSpatialMultiomics2024}, melanoma states\cite{minnoyeCrossspeciesAnalysisEnhancer2020, atakInterpretationAllelespecificChromatin2021}, human brain development\cite{mannensChromatinAccessibilityHuman2024}, mammalian and chicken brain cell types\cite{heckerEnhancerdrivenCellType2025} and pluripotent cell stem cells\cite{avsecBaseresolutionModelsTranscriptionfactor2021}. Furthermore, it has been demonstrated recently that the code learned by these models is sufficient to design novel enhancers with cell type-specific activity\cite{taskiranCelltypedirectedDesignSynthetic2024, dealmeidaTargetedDesignSynthetic2024, gosaiMachineguidedDesignCelltypetargeting2024, yinIterativeDeepLearningdesign2024}.\par

Organoids are a reductionistic biological model system that can be used to interrogate how cell types are encoded in the
genome. Organoids derived from pluripotent stem cells are especially interesting in this context, given that they
can be used to study how a genome can generate multiple distinct cell types\cite{mccauleyPluripotentStemCellderived2017,
takebeOrganoidsDesign2019, hoferEngineeringOrganoids2021}. Organoids can furthermore give insights into how genomes
interpret both cell intrinsic and extrinsic signals. For example, epigentic
modifications\cite{zenkSinglecellEpigenomicReconstruction2024}, transcription factor (TF)
perturbations\cite{fleckInferringPerturbingCell2023} or morphogen
signaling\cite{sanchis-callejaDecodingMorphogenPatterning2024}. Another benefit of organoids is that they can be used to
model biological systems that are otherwise - due to ethical and/or practical reasons - inaccessible for study. For
example, early human development.
One such system is early human development, the study of which is especially important for our understanding of human health and disease.
\par

@@

Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols and how they compare to their \textit{in vivo} counterparts at a transcriptomic level in for example, brain organoids\cite{heIntegratedTranscriptomicCell2024} and endoderm-derived organoids\cite{xuIntegratedTranscriptomicCell2023}. However, such comparisons at the level of the enhancer-code are still limited\cite{zenkSinglecellEpigenomicReconstruction2024, fleckInferringPerturbingCell2023}. In this study, we address this gap by determining the enhancer-codes of cell types/states generated by human neural tube organoids, a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021} and compare these codes to those of cell types/states of the head of a four weeks post conception human embryo.\par

During vertebrate development, both the neural tube and neural crest originate from the neural ectoderm. The neural tube
consists of proliferating neuronal progenitors that, through opposing gradients of SHH and BMP/Wnt signaling, achieve different identities along the dorso-ventral (DV) axis of the embryo. These early identities later give rise to distinct neuronal subtypes of the central nervous system\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020}. Neural crest cells originate from the neural plate border, through intermediate signals of BMP and Wnt\cite{simoes-costaEstablishingNeuralCrest2015}, and migrate throughout the body giving rise to a multitude of cell types including the peripheral nervous system, craniofacial skeleton and melanocytes\cite{simoes-costaEstablishingNeuralCrest2015}. The gene regulatory networks underlying both DV patterning of neuronal progenitors\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020} and neural crest induction, specification, migration and differentiation\cite{simoes-costaEstablishingNeuralCrest2015, candido-ferreiraMultilayeredTranscriptionalControl2023} are already well studied. Furthermore, studies using chromatin accessibility profiling methods at the bulk and single cell level, profiling enhancers encoding cell types and states of neuronal progenitors and neural crest cells are starting to emerge\cite{delasDevelopmentalCellFate2023, zhangCisregulatoryLogicIntegrating2024, huEndothelialRegulatoryCircuits2025} but detailed descriptions of the enhancer-code of individual cell types are still limited.\par

@@

We generated human neural tube organoids (hNTOs) to model early human neuronal and neural crest development. To this

@@

neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. To characterize the organoid-derived cell types to a greater level of detail at both

@@

Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) and cell
types missing from the organoid include endothelial cells, perivascular macrophages (PVM) / microglia (MIC) and otic placode (\textbf{Fig. 1c}).\par

@@

(\textit{SIX3}) or anterior spinal cord (\textit{HOXB3}) identity, in the embryo we observed a gradient of anterior-to-posterior identities (REF). Marked by the expression of \textit{SIX3}, \textit{EMX2}, \textit{FOXG1}, \textit{BARHL1}, \textit{EN1 }and \textit{HOXB3} (\textbf{Fig. 1g}).\par

@@

"$DeepNeuralTube^o$" and "$DeepNeuralTube^e$". The models use DNA sequence of genomic regions as input and

@@

train $DeepNeuralTube^o$) and embryo cells (to train $DeepNeuralTube^e$). To validate the ability of the models to

@@

(auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2b}; and

@@

enhancers with an auROC of 0.78 and 0.79 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$

@@

Next, we assessed the explainability of the models. Concretely we tested whether the model learned known

@@

TF-MoDISco (\textbf{REF}) to find \textit{de novo} motifs. For both known motifs and \textit{de novo}

@@

(\textbf{Fig. 2e}) regions with the exception of cluster 13 that is organoid specific (AP1 motif; \textbf{Supplementary Table S2}) and cluster

@@

Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ use known-motifs as features to classify regions based

@@

$DeepNeuralTube^o$ and $DeepNeuralTube^e$ explain this region similarly (Pearson correlation of x) and

@@

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_2.png}
  \caption{
    \textbf{DeepNeuralTube: an accurate sequence-to-function model revealing neuro-developmental enhancer code.} \\
    \textbf{a}, Schematic. \textbf{b} receiver-operator curves for identifying neural tube (left) and facial mesenchyme the
    $DeepNeuralTube^o$ and $DeepNeuralTube^e$. \textbf{c-e}, UMAP of TomTom [REF] motif-to-motif similarities
    using both \textit{de novo} motifs from DeepNeuralTube and known enriched motifs. Each dot is a motif. Motifs are
    colored by: cluster identity (\textbf{c}), model of origin (Continues next page)
  }
\end{figure}
\afterpage{\clearpage}

\begin{figure}[b]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) (\textbf{d}; only showing \textit{de novo} motifs in this case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
 \textbf{f}, Example representative motifs from clusters in b, showing motifs from motif enrichment (left) and found
    \textit{de novo} (right). \textbf{g}, contribution score for reference and mutant version of Vista enhancer hs111
    for $DeepNeuralTube^o$ (top two panels) and $DeepNeuralTube^e$ (bottom two panels) and enhancer-reporter
    assay for reference and mutant version of hs111 (right two panels) [REF]. \textbf{h}, contribution score for
    $DeepNeuralTube^o$ (top) and $DeepNeuralTube^e$ (bottom) of a genomic enhancer human neural crest
    enhancer (chr1:161200798-161201298; hg38) and enhancer-reporter assay in chicken embryo (HH4; right). Green is enhancer signal, pink is
    constitutive enhancer that serves as electroporation control.
  }
\end{figure}

\subsection*{
  The enhancer code underlying dorso-ventral patterning of neuronal progenitors
}
Focusing on neuronal progenitors, we identified topics representing cells and regions of different dorso-ventral
identities in both the organoid and the embryo data (\textbf{Fig. 3a}). From ventral to dorsal this is topic 36, 38, 33,
54 and 48 in the organoid data and topic 34, 38, 79, 88 and 58 in the embryo data. As an example we first investigated
two enhancers of \textit{SHH} that are strongly conserved and specific to the floorplate [\textbf{REF}]; SFPE1 and SFPE2.
\textit{SHH} is most strongly expressed in the most ventral topic (33 and 34 resp. in the organoids and the embryo; \textbf{Fig. 3b}).
SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells
belonging to the ventral topic, in both the organoid and embryo data (\textbf{Fig. 3c}). Explaining the accessibility of both
enhancers using DeepNeuralTube reveals two FOX (one partial) and two RFX (one partial) binding sites in SFPE1
and two FOX, a ZEB and a TEAD binding site in SFPE2 (\textbf{Fig. 3d}).\par

To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we calculated
contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube^o$ and $DeepNeuralTube^e$. We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities. Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3e}). Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns
underlying the different dorso-ventral states and the importance of the instances of these patterns is highly correlated across both
models (0.75 average Pearson correlation; \textbf{Fig. 3e}). For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD
and RFX instances is important (\textbf{Fig. 3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo cells (\textbf{Fig. 3g} and Supplementary Fig. S3). In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3e- g)}); PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3). For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), SOX instances are important which correlates with \textit{SOX2/3} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3). Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e} and Supplementary Fig. S3). Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig. 3h}). FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig. 3h}).\par

It has been reported that FOXA1/2 is a pioneer factor (REF). Indeed, when centering on FOX instances in both the organoid and
embryo data we observe a strong footprint when looking at Tn5 cut-sites (\textbf{Fig. 3i}). We then
asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this
TF and regions, bound by FOXA2, that are specific to the floorplate. For this purpose, we downloaded FOXA2 ChIP-seq
peaks in HepG2 and A549 and intersected these with floorplate peaks containing FOX instances. The floorplate peaks
that overlap with both ChIP-seq datasets we classified as regions that are generally bound by FOXA2 and those that
did not we classified as floorplate specific. Next, we quantified both the number of FOX instances per region and the
affinity of each individual instance (making use of the FOX motif score calculated using
FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) hypothesizing that the overall affinity of regions that are generally
bound by FOXA2 is higher than floorplate specific regions. Indeed, FOXA2 target regions that are specific to the floorplate mostly
have a single FOX instance while FOXA2 target regions that are general tend to have more instances (\textbf{Fig. 3j}).
Furthermore, in case there is a single FOX instance in a general FOXA2 target region the affinity of this instance tends
to be higher (\textbf{Fig. 3k}). Generating a \textit{de novo} motif from general and specific FOXA2 target regions that
have a single instance, we observe that FOX instances in regions that are general tend to be of the form
\textit{TGTTT\underline{AC}} while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).\par

@@

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38. SFPE2 and SFPE1 is highlighted. \textbf{d}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). \textbf{e}, Code table showing average contribution score of pattern instances for each dorso-ventral topic. Contribution scores for $DeepNeuralTube^o$ are not outlined. Contribution scores for $DeepNeuralTube^e$ are outlined and negated. \textbf{f}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. \textbf{g}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. \textbf{h}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. \textbf{i}, Scaled average and normalized number of cut sites (black, top) and contribution score (red, top) across genomic regions with a FOXA2 instance and heatmap showing normalized number of cut sites (bottom) for individual regions with a FOXA2 instance. Signal is centered on the FOXA2 instance start position. \textbf{j}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIp-seq signal (general; bottom) and those regions that don't overlap (specific; top). \textbf{k}, Boxplot (ADD NUMEBRS) showing $-log_10(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grantFIMOScanningOccurrences2011,schreiberTangermemeBiologicalSequence2025} for general regions (bottom) and specific regions (top) stratified by the number of FOXA2 instance per region. \textbf{l}, \textit{de novo} motif for regions with a single FOXA2 instance that are specific (left, top) or general (right, top) and heatmap color coded by nucleotide identity for individual regions for both specific (left, bottom) and general (right, bottom) regions that have a single FOXA2 instance.
  }
\end{figure}

@@

Given that we observed cell states from different AP domains in the neuronal progenitors from the embryo (but not
organoid; see \textbf{Fig. 1}) we decided to study the gene regulatory code underlying this patterning further. We
identified seven topics corresponding to different AP domains, from anterior to posterior this is Topic 61, Topic 59,
Topic 31, Topic 62, Topic 70, Topic 52 and Topic 71 (\textbf{Fig. 4a}). We first asked if the AP-program is distinct
from the DV program. In other words, is AP patterning encoded in the same or in different genomic regions as DV patterning? For this purpose, we calculated the pairwise correlation between region-topic
probabilities of all AP and DV topics (\textbf{Fig. 4b}). Both AP and DV patterning as well
as each of their sub-domains are mostly encoded in distinct genomic regions, with the exception of AP Topic 59 and 31
that correlate with the region-topic probability of DV Topic 38 (which corresponds to cells expressing \textit{NKX2-1/2}; \textbf{Fig. 4b})\par
